---
title: "Practical Machine Learning Project"
output: html_document
---

## Executive Summary
The goal of this project is to predict the manner in which the participants did the exercise. This is the "classe" variable in the training set. We will use the randomForest method in this analysis and cross validate. We will also estimate the out-of-sample error and test our model with the test dataset.

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Loading necesary packages

We will need the following packages to perform this analysis.

```{r packages}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(ggplot2)
library(corrplot)
```

## Downloading & reading data into R

Training & testing data can be found on the following link. Read data into R with the read.csv command.

``` {r data}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
train <- read.csv("./data/pml-training.csv")
test  <- read.csv("./data/pml-testing.csv")
```

## Cleaning the data

First off, we remove columns that have many missing values.

```{r cleaning}
train <- train[, colSums(is.na(train)) == 0] 
test  <- test[, colSums(is.na(test)) == 0] 
```

Next, we remove the columns that are not related to our analysis of accelerometers. These are the X, timestamp, and the window variables.

```{r cleaning2}
train <- train[, !(grepl("^X|timestamp|window", names(train)))]
test  <- test[, !(grepl("^X|timestamp|window", names(test)))]
```

We remove the rest of the variables that are meaningless to our analysis. Examples include variables that are missing in the test data or variables that have observations such as "DIV/0". We also remove the index variable for the same reason.

```{r cleaning3}
classe <- train$classe
train  <- train[, sapply(train, is.numeric)]
train$classe <- classe
test   <- test[, sapply(test, is.numeric)]
test   <- test[, -length(names(test))]
```

## Partitioning the data

We split the training data into subsets of training and test data for cross validation.

``` {r partition}
inTrain   <- createDataPartition(train$classe, p=0.75, list=F)
subtrain <- train[inTrain, ]
subtest  <- train[-inTrain, ]
```

## Fitting a model

Now that we have a training and test data set that we can use for cross validation, we can proceed to fitting a model. We will use the randomForest method because it automatically selects important variables. We will use the 5-fold cross validation in applying the model.

``` {r rf}
fitRF <- train(classe ~ ., data=subtrain, method="rf", trControl=trainControl(method="cv", 5), ntree=250)
fitRF
```

Now we can fit the model and test its performance on the validation test set.

```{r validation}
predictRF <- predict(fitRF, subtest)
confusionMatrix(subtest$classe, predictRF)
```

We observe that the accuracy is high at 99.45%. Based upon this result, we can estimate the out-of-sample error also.

```{r oose}
oose <- 1 - as.numeric(confusionMatrix(subtest$classe, predictRF)$overall[1])
oose
```

The estimated out-of-sample error is 0.55%.

## Test data prediction

We are now ready to plug in the real test data into our model.

```{r test}
predict(fitRF, test)
```

## Appendix

Included in the appendix are a couple plots to help visualization of our analysis.

* Correleation Matrix
```{r corrplot}
corrPlot <- cor(subtrain[, -length(names(subtrain))])
corrplot(corrPlot, method="color")
```

* Tree Visualization
```{r tree}
treeModel <- rpart(classe ~ ., data=subtrain, method="class")
prp(treeModel)
```